Syntax-Based Concept Alignment for Domain-Specific Machine Translation

### 2021-05-06 paper/paper.tex
finished general part of methodology, including diagram

@@

An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
In this paper, however, we focus on CE, and restrict its application to bilingual corpora.

@@

This work is a new attempt in the same direction, enabled by two interlingually oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{TODO:} and Grammatical Framework (GF) \cite{TODO:}.

@@

UD alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a ready-to-use, morphologically aware translation lexicon. This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiment using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{TODO:}. 
This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{TODO:} and \cite{TODO:}.

% system overview

>>>>

\usepackage{graphicx}

@@

% potentially move to future work directly
An interesting and less studied variant of CA is \textit{Concept Propagation} (CP), useful for cases where a set of concepts is already known and the goal is to identify the expressions corresponding to each of them in a new language, potentially even working with a different text in the same domain.
While our system implements basic CP functionalities, however, in this paper we focus on CE and restrict its application to bilingual corpora.

@@

This work is a new attempt in the same direction, enabled by two multilinguality-oriented grammar formalisms developed over the course of the last 25 years: Universal Dependencies (UD) \cite{TODO:} and Grammatical Framework (GF) \cite{TODO:}.

@@

UD parsing alone would then be sufficient to extract (or propagate) tree-to-tree alignments, but not to automate the generation of a ready-to-use, morphologically-aware translation lexicon. This is where GF comes into play: after correspondences are inferred from a parallel text, our proposed system is able to convert them to GF grammar rules, easy to embed in a domain-specific grammar but also making it possible to immediately evaluate the system by carrying out small-scale translation experiment using pre-existing grammatical constructions implemented in GF's Resource Grammar Library (RGL) \cite{TODO:}. 
This is made possible by \texttt{gf-ud}, a conversion tool described in \cite{TODO:} and \cite{TODO:}.

Concretely, the system we propose requires the following elements, whose reciprocal relations are shown in Figure \ref{overview}: \smallskip

\begin{itemize}
  \item a UD parser
  \item an alignment module based on dependency tree comparison
  \item a program, based on \texttt{gf-ud}, that converts the alignments into GF grammar rules and uses them to construct a translation lexicon
\end{itemize} \smallskip

\begin{figure}[h]
%  \centering
  \includegraphics[width=0.45\textwidth]{figures/overview.png}
  \caption[System overview]{System overview. Parts in grey are currently at the prototype stage and will not be further discussed in this paper.} \label{elems}
%  \label{overview}
\end{figure}
